Parallel Adaptive Importance Sampling

### 2018-01-25 response_JUQ.tex

\comment{However, I have some concerns.

1) The starting-point algorithm in Table 1, is very similar to the basic scheme suggested in, for instance,

V. Elvira, L. Martino, D. Luengo, M. F. Bugallo, Improving Population Monte Carlo: Alternative Weighting and Resampling Schemes, Signal Processing Vol. 131, pp. 77-91, 2017.

In practice, you use a deterministic approach for the IS weights. 
I believe that the authors have analyzed and studied this idea independently, but this similarity should be discussed in the introduction, at least. 
You can also point out that you also propose some different resampling or adaptation schemes that are not contained in the previous work.}{We agree that there are similarities to this other work, and we thank the referee for their comment. Additional references have been added, along with more detailed discussions about the novel aspects of this manuscript in the introduction and discussion.}

\comment{2) I do not understand why you use the word "parallel" in the name of the algorithm. In Algorithm 1, it is difficult to recognize the parallelization. The denominator in the weight that you employ, in fact, is a bottleneck for a possible parallelization. Please clarify this point (even in the text).}{We have added clarifications on this point to Section 3. The referee should note that in our convergence plots, the reason that we compare the algorithms with respect to the number of likelihood evaluations is that we envisaged that this method would be of most use when the likelihood is expensive, for example when it requires a high resolution approximation of the solution of a PDE, and/or there is a very large amount of data to be taken into account. In this context, it would be useful to spread the work of computing the likelihood for the ensemble members across a cluster, hence our use of the word ``parallel''. We acknowledge and thank the referee for indicating the issues with the denominator. In the case where the likelihood is expensive, these overhead costs, along with that of the resampler, should no longer be the bottleneck. This was not adequately explained in the text, and as such we have added a further discussion in Section 3.

@@

% We agree with this comment from the referee, as it may not have been made clear enough. With an ensemble of states, it is possible to distribute the computation of the weights across a set of cores. We have added a comment on this to Section ****. However, as discussed in the response to the first referee, we ourselves implemented the algorithm serially. Therefore, we have decided to change the name of the algorithm to reflect this, to ``Optimal Transport Adaptive Importance Sampling'' or OTAIS. Changes throughout the manuscript have been made to reflect this.
}

\comment{3) Please, add a table of acronyms that can help the reader. Now it is a bit confusing.}{We have added a glossary of acronyms to an appendix.} %half done -  complete

@@

\comment{4) The references should be completed with other important contributions (considering also the same kind of weights, in some cases):

- O.Cappe, R.Douc, A.Guillin, J.M.Marin, C.P.Robert, Adaptive importance sampling in general mixture classes, Statistics and Computing 18 (2008) 447-459. 
- R. Douc, A. Guillin, J. M. Marin, C. P. Robert, Convergence of adaptive mixtures of importance sampling schemes, Annals of Statistics 35 (2007) 420-448. 
- R. Douc, A. Guillin, J. M. Marin, C. P. Robert, Minimum variance importance sampling via population Monte Carlo, ESAIM: Probability and Statistics 11 (2007) 427-447. 
- L. Martino, V. Elvira, D. Luengo, J. Corander, Layered Adaptive Importance Sampling, Statistics and Computing, Vol. 27 (3), pp. 599-623, 2017.}{We thank the referee for thier suggestions for other highly relevant references.} %done

\section*{Associate Editor comments}
\comment{A very similar version of the authors’ PAIS algorithm has been studied before, as explained by the second referee. Moreover, I could argue that the PAIS algorithm is a particular case since weights are considered deter- ministic. This is a serious issue in the paper, let alone that the authors are unaware of the series of papers mentioned by referee 2, that are closely related to their work.
This should be explained in detail and the paper original perspective and contributions be further highlighted.}{Following suggestions by the second referee and the associate editor, we have added a significant new section to the introduction to highlight the other ensemble importance schemes already in the literature. We have also highlighted further the contribution made by this manuscript in the introduction and discussion sections.}%done

\comment{An other serious concern is why not use PMC-MCMC for the updating scheme in Algorithm 1. That would be as costly (since it also involves evaluating the posterior) but the yis will have the correct distribution.}{This comment is not clear to us - what is the ``updating scheme'' in Algorithm 1?? Also, the (weighted) $y_i$ are distributed according to the posterior, as discussed in Section 6. We'd be happy to address this point if the AE is able to clarify it for us.}%nothing to do

\comment{The ”automatic” tuning will be as useful as the what traditional adaptive tuning is: in the limit, the proposal cov matrix converges to the cov matrix of the target. This is ok for Gaussian type targets but will fail for multimodal, multiscale distributions, mostly present in non-trivial forward maps, as those suggested in section 2. Something similar happens with gradient based proposals, since from the onset they are not conceived to scape local maxima.}{This is an interesting point, and the subject of our most recent paper in this area, in which transport maps are used to stabilise and accelerate the PAIS methodology, and also addressing exactly the point raised by the AE. We have added a discussion on this topic in the discussion section, proposing it as an area for future study (our new manuscript will be completed in the next few weeks).}%done

\comment{In Algorithm 1, note that in step 5 the actual number of different particles could be drastically reduced and in some cases collapsed to 1. As mentioned by the authors, this is a basic concern in SIR type algorithms, that they may become totally inefficient. Then, why this seems not to bother the authors and no provision was included in their algorithm to mend this problem?}{Provision is made within the code for such a catastrophic collapse of the weights, and it was remiss of us not to include details on this. We have added details on this to the end of Section 3.}%done

\comment{The Radon-Nokodyn version of Bayes theorem in (2.3) is unnecessary in the actual finite dimensional context the authors work with. The authors should state that, although there is an infinite dimensional version for the posterior distribution, their discussion is restricted to the parametric, finite dimensional case.}{We thank the AE for this comment. There is no need for this version of Bayes' theorem here, and so we have switched to the finite dimensional version.}%done

\comment{Note that in (2.4) the total measure of the space for ηˆ is M (!).}{Thanks to the AE for picking up on this typo, this has now been fixed.}%done

### 2018-01-25 PAIS.tex
Referee 2 and AE comments addressed

Ensemble importance sampling schemes also exist, e.g. population Monte Carlo (PMC)~\cite{cappe2012population}. PMC uses an ensemble to build a mixture or kernel density estimate (KDE) of the
posterior distribution. The efficiency of
this optimisation is restricted by the component kernel(s) chosen, and
the quality of the current sample from the posterior. Extensions, such
as adaptive multiple importance sampling algorithm
(AMIS)~\cite{cornuet2012adaptive}, and adaptive population importance sampling
(APIS)~\cite{martino2015adaptive} have enabled these methods to be
applied to various applications, including population
genetics~\cite{siren2011reconstructing}.

@@

@@

As discussed in \cite{stuart2010inverse,cotter2009bayesian},
in order for this inverse problem to be well-posed in the Bayesian
sense, we require the posterior distribution, $\mu_Y$, to be absolutely
continuous with respect to the prior, $\mu_0$. A
minimal regularity prior can be chosen informed by regularity results
of the observational operator $\mathcal{G}$. Given such a prior, then
the Radon-Nikodym derivative of the posterior measure, $\mu_Y$, with
respect to the prior measure, $\mu_0$, is proportional to the
likelihood:
\begin{equation}\label{eqn:RND}
	\frac{d\mu_Y}{d\mu_0} \propto \exp \left ( -\Phi(x^*) \right ).
\end{equation}

@@

In this work we use these particle filters as resampling methods. We
approximate the posterior distribution, $\mu_Y$, with a small sample
of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. This distribution, $\hat{\mu}_Y
\approx \mu_Y$, can then be resampled, using the ETPF or otherwise,
into a new distribution, $\hat{\eta} \approx \mu_Y$,
\begin{equation}\label{eqn:resampler}
	\hat{\mu}_Y = \sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 n_i\delta_{x_i}(\cdot) = \hat{\eta}.
\end{equation}
where $n_i \in \{0, 1, \dots,M\}$ and $\sum_{i=1}^M \! n_i = M$. This
new sample $\{x_i\}_{i=1}^M$ is equally weighted.

@@

In each iteration of the algorithm, we are required to compute the
likelihood for each member of the ensemble. In the case where the
likelihood is expensive to compute, for example because it requires
the numerical approximation of an ODE or PDE, and/or there is a very
large amount of observations in the data, it will be expedient to
split this computational effort across multiple cores. Whether this is
efficient will depend on the architecture of the machine, and the
relative cost of communication across the cores, to the cost of the
likelihood evaluation. Evaluation of the denominator in the weights
also contributes to the overhead costs of this approach over and above
a standard MCMC method, but this too can be parallelised efficiently,
depending on communication costs of the architecture of the machine.

@@

@@

We have demonstrated that the framework that we have considered, with
the use of state-of-the-art optimal transport-based resampling, can
reduce the number of likelihood evaluations required to characterise
complex posterior distributions in low dimensions to a given
degree. We have also introduced a greedy approximation to this
resampler, which drastically reduces the cost, at the loss of some
accuracy, which can be mitigated with the use of larger ensembles.

>>>>

Ensemble importance sampling schemes also exist, e.g. population Monte Carlo (PMC)~\cite{cappe2012population}. PMC uses an ensemble to build a mixture or kernel density estimate (KDE) of the
posterior distribution. The efficiency of
this optimisation is restricted by the component kernel(s) chosen, and
the quality of the current sample from the posterior\cite{cappe2008adaptive,douc2007convergence,douc2007minimum}. Extensions, such
as layered adaptive importance sampling\cite{martino2017layered}, adaptive multiple importance sampling algorithm
(AMIS)~\cite{cornuet2012adaptive}, and adaptive population importance sampling
(APIS)~\cite{martino2015adaptive} have enabled these methods to be
applied to various applications, including population
genetics~\cite{siren2011reconstructing}.

@@

The method considered is similar to population Monte Carlo methods
which have been previously studied, for example in
\cite{elvira2017improving}. However our approach exploits the current
state-of-the-art resamplers. Higher accuracy resamplers, that provide
a deterministic resample which optimally reflects the statistics of
the original sample, lead to better mixture approximations of the
posterior. These approximations can then be used to construct stable
and efficient importance sampling proposals. The cost of the
state-of-art resamplers can be prohibitive for large ensemble sizes,
which are necessary for sampling in moderately higher dimensions, and
as such we also present a greedy approximation of the optimal
transport resampler, which provides us with good results for a
fraction of the cost. We also detail how scaling parameters in the
MCMC proposals can be quickly tuned, leading us to an efficient and
fully-automated algorithm.

@@

Then, according to Bayes' theorem, the posterior distribution that we
are interested in characterising is given, up to a constant of proportionality, by
\begin{equation*}
	\pi_Y \propto \exp \left ( -\Phi(x^*) \right )\pi_0(x^*),
\end{equation*}
where $\pi_0$ is the prior density on the quantity of interest $x^*$.

% As discussed in \cite{stuart2010inverse,cotter2009bayesian},
% in order for this inverse problem to be well-posed in the Bayesian
% sense, we require the posterior distribution, $\mu_Y$, to be absolutely
% continuous with respect to the prior, $\mu_0$. A
% minimal regularity prior can be chosen informed by regularity results
% of the observational operator $\mathcal{G}$. Given such a prior, then
% the Radon-Nikodym derivative of the posterior measure, $\mu_Y$, with
% respect to the prior measure, $\mu_0$, is proportional to the
% likelihood:
% \begin{equation}\label{eqn:RND}
% 	\frac{d\mu_Y}{d\mu_0} \propto \exp \left ( -\Phi(x^*) \right ).
% \end{equation}

% %%%%%%

@@

In this work we use these particle filters as resampling methods. We
approximate the posterior distribution, $\mu_Y$, with a small sample
of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. This distribution, $\hat{\mu}_Y
\approx \mu_Y$, can then be resampled, using the ETPF or otherwise,
into a new distribution, $\hat{\eta} \approx \mu_Y$,
\begin{equation}\label{eqn:resampler}
	\hat{\mu}_Y = \sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot) = \hat{\eta}.
\end{equation}
where $n_i \in \{0, 1, \dots,M\}$. This
new sample $\{x_i\}_{i=1}^M$ is equally weighted.

@@

In each iteration of the algorithm, we are required to compute the
likelihood for each member of the ensemble. In the case where the
likelihood is expensive to compute, for example because it requires
the numerical approximation of an ODE or PDE, and/or there is a very
large amount of observations in the data, it will be expedient to
split this computational effort across multiple cores. Whether this is
efficient will depend on the architecture of the machine, and the
relative cost of communication across the cores, to the cost of the
likelihood evaluation. Evaluation of the denominator in the weights
also contributes to the overhead costs of this approach over and above
a standard MCMC method, but this too can be parallelised efficiently,
depending on communication costs of the architecture of the machine,
reducing the $\mathcal{O}(M^2)$ cost to $\mathcal{O}(M)$ over $M$
cores, for example.

@@

It is also possible, especially during the early stages of the
algorithm, where the mixture proposal is a poor approximation of the
target distribution, for a sample to be produced with a weight which
is orders of magnitude bigger than the rest. This is a problem, since
the resampling step will then lead to a vector of particles all
centred at the outlier sample. Fortunately these spikes in weights can
easily be detected, and in such a case, the sample for this iteration
can be removed. The ETPF can then be used to formulate a new mixture
sample with one particle placed at the outlier point, and the others
distributed according to the particle positions before the problematic
iteration. This ensures that the region where the problem occurred is
better represented by the mixture, and prevents further spikes in this region.

@@

We have demonstrated that the framework that we have considered, with
the use of state-of-the-art optimal transport-based resampling, can
reduce the number of likelihood evaluations required to characterise
complex posterior distributions in low dimensions to a given
degree. We have also introduced a greedy approximation to this
resampler, which drastically reduces the cost, at the loss of some
accuracy, which can be mitigated with the use of larger
ensembles. Lastly, we have detailed how scaling parameters in the MCMC
proposals that are used within the mixture distribution can be quickly
and efficiently
tuned in an automated way. However, tuning the variances and
covariances of the mixture proposal components
globally is likely to be of limited use for multimodal problems where
the modes have very different covariances, or where there is curved
ridges in the density. There are a variety of possible solutions to
this problem that would be worthy of future consideration, not least
the potential to use transport maps\cite{el2012bayesian,parno2014transport} to map Gaussian mixtures to
highly complex non-Gaussian approximations of the posterior. Such a
map could encode local covariance information, and lead to accelerated
and stable PAIS sampling. This is an interesting area for future study.

\begin{appendix}

\section{Glossary of acronyms}

MCMC
RWMH
MALA
AIS
PMC
PAIS
ETPF
AMR
PAIS-RW
PAIS-MALA
APAIS-RW
APAIS-MALA

\end{appendix}
